\documentclass[11pt,a4paper]{article}

% -----------------------------------------------------------------------
% Packages
% -----------------------------------------------------------------------
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{geometry}
\geometry{margin=2.5cm}

% -----------------------------------------------------------------------
% Theorem environments
% -----------------------------------------------------------------------
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

% -----------------------------------------------------------------------
% Macros
% -----------------------------------------------------------------------
\newcommand{\ELBO}{\mathcal{L}_{\mathrm{VDT}}}
\newcommand{\KL}[2]{D_{\mathrm{KL}}\!\left(#1\,\|\,#2\right)}
\newcommand{\E}[1]{\mathbb{E}\!\left[#1\right]}
\newcommand{\Lmat}{\mathbf{L}}
\newcommand{\Umat}{\mathbf{U}}
\newcommand{\Smat}{\mathbf{S}}
\newcommand{\Wmat}{\mathbf{W}}
\newcommand{\Omegamat}{\boldsymbol{\Omega}}
\newcommand{\lam}{\boldsymbol{\lambda}}
\newcommand{\normal}{\mathcal{N}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{^{\intercal}}

% -----------------------------------------------------------------------
% Metadata
% -----------------------------------------------------------------------
\title{%
  \textbf{Vibrational Deduction Transformer (VDT)}\\[4pt]
  \large A Spectral-PPCA Variational Autoencoder with Graph-Wiring
  Latent Space and Associative Spectral Memory}

\author{%
  Lorenzo Moriondo\\[2pt]
\small Independent Researcher --- tuned.org.uk\\
\small ORCID: 0000-0002-8804-2963\\
  \href{https://github.com/tuned-org-uk/vibrational-deduction-transfomer}{%
    github.com/tuned-org-uk/vibrational-deduction-transfomer}}

\date{June 2026 (draft)}

% -----------------------------------------------------------------------
\begin{document}
% -----------------------------------------------------------------------

\maketitle

\begin{abstract}
We introduce the \emph{Vibrational Deduction Transformer} (VDT), a generative
architecture that treats the latent space of a Variational Autoencoder (VAE) as
a vibrational system grounded in Rayleigh's \emph{Theory of Sound}.
The VDT decoder maps a stochastic latent code $\mathbf{z}$ to a
\emph{spectral loading matrix} $\Wmat = \Umat_{1:q}\,\mathrm{diag}(\boldsymbol{\omega})\,\Smat$,
where $\Umat_{1:q}$ are the $q$ lowest-frequency eigenvectors of a frozen
ArrowSpace graph-Laplacian index $\Lmat(\mathcal{I})$, $\Smat$ is a learnable
coefficient matrix in that eigenbasis, and $\boldsymbol{\omega}$ are mode weights
drawn from a tau-mode variational prior.
This loading matrix parametrises a differentiable Laplacian $\Lmat(\mathbf{z})$
over which a heat-kernel \emph{tau-mode diffusion} decoder reconstructs the input.
The objective is a three-term Evidence Lower Bound (ELBO) comprising a
Gaussian reconstruction term, an isotropic latent KL, an eigenvalue-weighted
spectral-basis KL penalising high-frequency loadings (spectral Occam's razor),
and a Gamma-vs-Exponential mode-frequency KL for spectral band selection.
After training, the model emits a \emph{spectral artefact}
$\mathcal{A}(\mathcal{I})=\{\hat{\Wmat}, \hat{\boldsymbol{\omega}}, \Smat_{\mathrm{mem}}\}$
that initialises a Hopfield-type associative memory for a downstream transformer.
Motivated by Aaronson's observation that quantum interference follows from signed
(real-valued positive-negative) amplitudes, the VDT avoids complex numbers:
interference is encoded in the sign structure of Laplacian edge weights rather
than complex phases.
Benchmarks on CORA, PubMed, and synthetic spring-network graphs demonstrate
entropy-controlled spectral graph generation and strong linear-probe accuracy
of frozen latent representations.
\end{abstract}

\tableofcontents
\newpage

% -----------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% -----------------------------------------------------------------------

Transformer-based generative models typically organise learned representations
in a flat isotropic Gaussian latent space, discarding the geometric structure
of the data.  When the data lives on a graph---citation networks, molecular
graphs, knowledge graphs, or physical spring networks---this structure carries
critical semantic information that a flat prior cannot capture.

The Vibrational Deduction Transformer (VDT) is motivated by three converging
lines of inquiry:

\begin{enumerate}
  \item \textbf{Rayleigh's vibrational analogy.}
        Rayleigh~\cite{rayleigh1877} showed that the natural oscillation modes
        of a continuous medium are the eigensolutions of its governing
        differential operator.  Graph Laplacian eigenvalues are the discrete
        analogue: small eigenvalues correspond to smooth, low-frequency modes
        (the ``standing wave'' state); large eigenvalues correspond to
        high-frequency, disordered modes (the ``rest'' or collapsed state).
        The VDT treats the \emph{low-frequency oscillation state as the
        low-entropy prior} and high-frequency disorder as high entropy,
        directly reversing the usual thermodynamic framing: the natural state
        of the string is to oscillate, not to be still.

  \item \textbf{Real-valued quantum amplitudes.}
        Aaronson~\cite{aaronson2013} observes in Chapter~9 of
        \emph{Quantum Computing Since Democritus} that the essential features
        of quantum probability---interference, entanglement structure,
        the Born rule---are recoverable from signed real amplitudes without
        complex numbers, provided the update rule is unitary.
        The VDT implements this insight: edge weights in $\Lmat(\mathbf{z})$
        can be negative (via the spectral loading parametrisation), producing
        destructive interference in the heat kernel without any complex
        arithmetic.

  \item \textbf{ArrowSpace / Graph Wiring.}
        The ArrowSpace computational framework~\cite{moriondo2026arrowspace}
        defines a differentiable Laplacian builder and a spectral cost
        ($J_{\text{freq}}$) that penalises high-frequency wiring.
        The VDT replaces this hard penalty with a \emph{variational}
        counterpart---a KL divergence over spectral loadings and mode
        weights---yielding a proper Bayesian ELBO.
\end{enumerate}

The paper is organised as follows.
Section~\ref{sec:background} reviews prerequisite material.
Section~\ref{sec:architecture} presents the full VDT architecture.
Section~\ref{sec:elbo} derives the three-term ELBO.
Section~\ref{sec:memory} describes the spectral artefact and associative memory.
Section~\ref{sec:stability} covers the stability analysis.
Section~\ref{sec:experiments} reports empirical results.
Section~\ref{sec:related} discusses related work.
Section~\ref{sec:conclusion} concludes.

% -----------------------------------------------------------------------
\section{Background}
\label{sec:background}
% -----------------------------------------------------------------------

\subsection{Graph Laplacians and Spectral Graph Theory}
\label{sec:background:laplacian}

Given an undirected weighted graph $G=(V,E,w)$ with $N=|V|$ nodes, the
\emph{combinatorial Laplacian} is $\Lmat = \mathbf{D} - \mathbf{A}$, where
$\mathbf{D}$ is the degree matrix and $\mathbf{A}$ the weighted adjacency.
The normalised Laplacian $\tilde{\Lmat} = \mathbf{D}^{-1/2}\Lmat\mathbf{D}^{-1/2}$
has eigenvalues $0 = \lambda_1 \leq \lambda_2 \leq \cdots \leq \lambda_N \leq 2$.
The eigenvectors $\{\mathbf{u}_k\}$ form an orthonormal basis (the \emph{graph Fourier
basis}), and the \emph{Rayleigh quotient}
\begin{equation}
  \mathcal{R}(\mathbf{v}) = \frac{\mathbf{v}\T\Lmat\mathbf{v}}{\mathbf{v}\T\mathbf{v}}
  \label{eq:rayleigh}
\end{equation}
gives the frequency (smoothness cost) of signal $\mathbf{v}$ on the graph~\cite{chung1997}.
This is the discrete analogue of Rayleigh's classical frequency functional for
continuous media~\cite{rayleigh1877}.

\subsection{Probabilistic PCA and Variational Autoencoders}
\label{sec:background:ppca}

Probabilistic PCA (PPCA)~\cite{tipping1999} provides a generative model
$p(\mathbf{x}|\mathbf{z}) = \normal(\mathbf{W}\mathbf{z}+\boldsymbol{\mu}, \sigma^2\mathbf{I})$
where $\Wmat \in \R^{D\times q}$ is the loading matrix and $\mathbf{z}\sim\normal(\mathbf{0},\mathbf{I})$.
Variational Autoencoders (VAEs)~\cite{kingma2013} extend this to nonlinear decoders
by maximising the Evidence Lower Bound
$\ELBO = \E{\log p(\mathbf{x}|\mathbf{z})} - \KL{q(\mathbf{z}|\mathbf{x})}{p(\mathbf{z})}$.
The reparameterisation trick $\mathbf{z} = \boldsymbol{\mu} + \boldsymbol{\sigma}\odot\boldsymbol{\epsilon}$,
$\boldsymbol{\epsilon}\sim\normal(\mathbf{0},\mathbf{I})$, enables gradient-based optimisation.

\subsection{Real Quantum Amplitudes and Signed Interference}
\label{sec:background:quantum}

Aaronson~\cite{aaronson2013} demonstrates that the observable consequences of
quantum mechanics---interference, entanglement structure, Born-rule probabilities---can
be derived from the assumption of signed real probability amplitudes, provided
the time-evolution operator is a \emph{real orthogonal} (rather than complex unitary)
matrix.  The key insight is:

\begin{quote}
  ``If we allow \emph{negative} probability amplitudes, then we get
  quantum-like interference without complex numbers.'' \cite{aaronson2013}
\end{quote}

The VDT operationalises this: the loading matrix $\Wmat$ has real-valued entries
that can be positive or negative, encoding interference in the Laplacian edge
structure.  The heat kernel $\mathbf{K}_\tau = \Umat\,\exp(-t\boldsymbol{\Lambda})\,\Umat\T$ is
a real symmetric positive-definite matrix and acts as the ``real unitary channel''.

% -----------------------------------------------------------------------
\section{Architecture}
\label{sec:architecture}
% -----------------------------------------------------------------------

\subsection{Overview}
\label{sec:architecture:overview}

The VDT is a \emph{Spectral-PPCA VAE} whose generative path is mediated by a
learned graph wiring constrained to the eigenbasis of an ArrowSpace index $\mathcal{I}$.
Figure~\ref{fig:arch} summarises the data flow.

\begin{figure}[ht]
  \centering
  \begin{verbatim}
Input x (B, D)              Embedding table E (N, D)
    |                                |
    +--[lambda-fingerprint L(I)]-----+
    |
    v
+----------------------------+
| WiringEncoder              |
|  MLP + lambda-fingerprint  |
|  --> z, mu, log_var,       |
|      log_a, log_b          |
+----------------------------+
    |                  |
  (z, mu, log_var) <-- reparameterise
  (log_a, log_b)   --> ModeWeightHead --> q(omega)
    |
  z_to_q: z (B, latent_dim) --> z_q (B, q)
    |
    v
+------------------------------------------+
| SpectralLoadingDecoder                   |
|  z_q, U_{1:q} --> W, omega, S, log_var_S |
|  W = U_{1:q} diag(omega) S               |
+------------------------------------------+
    |                   |
  W --> DifferentiableLaplacian --> L(z)    kl_S
    |
    v
+------------------+     +---------+
| DiffusionDecoder | <-- |    E    |
| TauModeDiffusion |     +---------+
+------------------+
    |
  x_hat (B, D) --> recon + kl_tau + kl_z + kl_S = loss
  \end{verbatim}
  \caption{VDT data flow. The index $\mathcal{I}$ enters only through the frozen
    eigenpair $(\Umat_{1:q}, \boldsymbol{\Lambda}_{1:q})$; no Laplacian is
    evaluated at training time.}
  \label{fig:arch}
\end{figure}

\subsection{WiringEncoder}
\label{sec:architecture:encoder}

The encoder is a multi-layer perceptron (MLP) augmented with a
$\lambda$-\emph{fingerprint}---a fixed scalar summary of the Laplacian spectrum
that encodes the global spectral geometry of the graph:

\begin{equation}
  \text{fingerprint}(\Lmat(\mathcal{I})) \;=\; f\bigl(\lambda_1,\ldots,\lambda_q\bigr)
  \in \R^{d_{\text{fp}}},
  \label{eq:fingerprint}
\end{equation}

where $f$ is a deterministic aggregation function (e.g.\ normalised eigenvalue
histogram).  The encoder produces:

\begin{equation}
  \mathbf{z},\,\boldsymbol{\mu},\,\log\boldsymbol{\sigma}^2,\,
  \log\mathbf{a},\,\log\mathbf{b}
  \;=\; \text{WiringEncoder}(\mathbf{x},\,\Lmat(\mathcal{I})),
  \label{eq:encoder}
\end{equation}

where $\mathbf{z}\in\R^{d_z}$ is the reparameterised VAE latent,
$(\mathbf{a},\mathbf{b})\in\R^q_{>0}\times\R^q_{>0}$ are the shape and rate
parameters of the Gamma variational posterior over mode weights.

\subsection{SpectralLoadingDecoder}
\label{sec:architecture:decoder}

A linear projection $\mathbf{z}_q = \mathbf{W}_{z\to q}\,\mathbf{z} \in\R^q$ bridges
the VAE latent dimension $d_z$ to the spectral mode dimension $q$.
The spectral loading decoder then produces:

\begin{align}
  \Smat &= \text{S\_net}(\mathbf{z}_q) \in \R^{q\times q},\quad
  \text{posterior mean of spectral coefficients},
  \label{eq:S} \\
  \log\boldsymbol{\sigma}^2_{\Smat} &= \text{log\_var\_S\_head}(\mathbf{z}_q)
  \in \R^{q\times q},\quad
  \text{\emph{independent} posterior log-variance},
  \label{eq:logvarS} \\
  \boldsymbol{\omega} &= \exp(\text{omega\_net}(\mathbf{z}_q)) \in \R^q_{>0},\quad
  \text{mode weights},
  \label{eq:omega} \\
  \Wmat &= \Umat_{1:q}\,\mathrm{diag}(\boldsymbol{\omega})\,\Smat
  \;\in\; \R^{d\times q}.
  \label{eq:W}
\end{align}

Equation~\eqref{eq:W} is the central structural commitment of the VDT:
the loading matrix lives in the Laplacian eigenbasis.
High-frequency directions are suppressed both by the mode weight prior
(Section~\ref{sec:elbo:kltau}) and by the eigenvalue-weighted prior on $\Smat$
(Section~\ref{sec:elbo:kls}).  The independence of $\log\boldsymbol{\sigma}^2_{\Smat}$
from $\Smat$ is a deliberate design decision (fix~\#52): the old proxy
$\log(\Smat^2+\epsilon)$ conflated posterior mean and variance, producing
biased KL gradients.

\subsection{DifferentiableLaplacian}
\label{sec:architecture:laplacian}

The loading matrix $\Wmat$ is converted to a graph Laplacian by:

\begin{equation}
  w_{ij}(\Wmat) = w_{ij}^{\text{base}} \cdot \sigma\!\left(-\|\Wmat_i - \Wmat_j\|^2\right),
  \quad
  \Lmat(\mathbf{z}) = \mathbf{D}(\Wmat) - \mathbf{A}(\Wmat),
  \label{eq:difflaplacian}
\end{equation}

where $\sigma$ is the logistic sigmoid and $w_{ij}^{\text{base}}$ is a frozen
base-graph weight.  The construction is fully differentiable through $\Wmat$.
Negative edge weights are permitted by this parametrisation whenever the
spectral loading produces $\|\Wmat_i-\Wmat_j\|^2 < 0$ after sigmoid, which
corresponds to anti-correlated spectral directions and encodes
\emph{destructive interference} in the heat kernel---the real-amplitude
counterpart of complex phase cancellation.

\subsection{DiffusionDecoder (TauModeDiffusion)}
\label{sec:architecture:diffusion}

The decoder applies a truncated heat kernel to the embedding table $\mathbf{E}$:

\begin{equation}
  \mathbf{K}_\tau = \Umat_{\tau}\,\exp(-\tau\,\boldsymbol{\Lambda}_{\tau})\,\Umat_{\tau}\T,
  \quad
  \hat{\mathbf{x}} = (\mathbf{K}_\tau\,\mathbf{E})_{[\text{node\_idx}]} + \text{MLP}(\mathbf{z}_q),
  \label{eq:diffusion}
\end{equation}

where the summation is truncated to the leading $\tau_{\text{modes}}$ eigenpairs
of $\Lmat(\mathbf{z})$.  The MLP refinement corrects residuals not captured by the
spectral path.  The heat kernel $\mathbf{K}_\tau$ is a real symmetric
positive-semi-definite operator---the real-valued analogue of a quantum
channel~\cite{aaronson2013,belkin2003}.

% -----------------------------------------------------------------------
\section{The Three-Term ELBO}
\label{sec:elbo}
% -----------------------------------------------------------------------

The VDT objective is the negative ELBO:

\begin{equation}
  \boxed{\mathcal{L}_{\mathrm{VDT}} =
    \underbrace{-\E{\log p(\mathbf{x}|\mathbf{z},\Wmat)}}_{\text{NLL recon}}
    + \underbrace{\KL{q(\mathbf{z})}{\normal(\mathbf{0},\mathbf{I})}}_{\text{latent KL}}
    + \underbrace{\KL{q(\Smat)}{p(\Smat|\mathcal{I})}}_{\text{spectral basis KL}}
    + \underbrace{\KL{q(\boldsymbol{\omega})}{p(\boldsymbol{\omega}|\tau,\boldsymbol{\Lambda})}}_{\text{tau-mode KL}}}
  \label{eq:elbo}
\end{equation}

All four terms are non-negative; minimising their sum is equivalent to
maximising the ELBO.  We derive each term below.

\subsection{Term 1: Reconstruction}
\label{sec:elbo:recon}

Under the Gaussian output model the reconstruction term is:

\begin{equation}
  -\E{\log p(\mathbf{x}|\mathbf{z},\Wmat)}
  = \frac{\|\mathbf{x} - \hat{\mathbf{x}}\|^2}{2\sigma^2}
  + \frac{D}{2}\log(2\pi\sigma^2).
  \label{eq:recon}
\end{equation}

In the implementation the constant $\frac{D}{2}\log(2\pi)$ is dropped;
$\sigma$ is a learnable scalar log-variance.

\subsection{Term 2: Isotropic Latent KL}
\label{sec:elbo:klz}

The standard VAE isotropic KL~\cite{kingma2013}:

\begin{equation}
  \KL{q(\mathbf{z}|\mathbf{x})}{\normal(\mathbf{0},\mathbf{I})}
  = -\frac{1}{2}\sum_{j=1}^{d_z}\left(1 + \log\sigma_j^2 - \mu_j^2 - \sigma_j^2\right).
  \label{eq:klz}
\end{equation}

\subsection{Term 3: Spectral-Basis KL}
\label{sec:elbo:kls}

The loading matrix $\Smat\in\R^{q\times q}$ carries a Gaussian prior shaped by
the frozen eigenvalues:

\begin{equation}
  p(\Smat|\mathcal{I}) \;\propto\;
  \exp\!\left(-\frac{\lambda_S}{2}\,\mathrm{tr}\!\left(\Smat\T\boldsymbol{\Lambda}_{1:q}\Smat\right)\right),
  \label{eq:priorS}
\end{equation}

where $\lambda_S > 0$ is a regularisation hyperparameter and
$\boldsymbol{\Lambda}_{1:q} = \mathrm{diag}(\lambda_1,\ldots,\lambda_q)$.
Large eigenvalues penalise high-frequency components more strongly---a
\emph{spectral Occam's razor}.  The posterior is
$q(\Smat)=\normal(\Smat_0,\,\mathrm{diag}(\exp(\log\boldsymbol{\sigma}^2_{\Smat})))$.
The KL is closed-form under the Gaussian-Gaussian conjugacy.

\subsection{Term 4: Tau-Mode Frequency KL}
\label{sec:elbo:kltau}

Mode weights $\omega_k>0$ receive an exponential prior indexed by frequency:

\begin{equation}
  p(\omega_k|\tau,\lambda_k) = \mathrm{Exponential}(\tau\lambda_k),
  \label{eq:prioromega}
\end{equation}

giving heavy support on low-frequency modes (small $\lambda_k$).
The variational posterior is a Gamma: $q(\omega_k)=\mathrm{Gamma}(a_k,b_k)$.
The KL between a Gamma and an Exponential is closed-form:

\begin{equation}
  \KL{\mathrm{Gamma}(a,b)}{\mathrm{Exp}(r)}
  = \log b - \log r + \ln\Gamma(a) + (1-a)\psi(a) + \frac{a}{b}r,
  \label{eq:kltau}
\end{equation}

where $\psi$ is the digamma function.  No Monte Carlo sampling is required.

\subsection{Comparison with the Base Wiring AE}
\label{sec:elbo:comparison}

\begin{table}[ht]
  \centering
  \caption{ELBO structure: Base Wiring AE vs.\ VDT.}
  \label{tab:elbo}
  \begin{tabular}{lll}
    \toprule
    \textbf{Term} & \textbf{Wiring AE (base)} & \textbf{VDT} \\
    \midrule
    Reconstruction   & $-\log p(\mathbf{x}|\mathbf{z})$         & Same \\
    Latent KL        & $\KL{q(\mathbf{z})}{\normal(\mathbf{0},\mathbf{I})}$ isotropic & Same \\
    Spectral basis   & None                          & $\KL{q(\Smat)}{p(\Smat|\mathcal{I})}$ eigenvalue-weighted \\
    Frequency        & $\alpha J_{\text{freq}}$ hard penalty & $\KL{q(\boldsymbol{\omega})}{\mathrm{Exp}(\tau\lambda_k)}$ variational \\
    \bottomrule
  \end{tabular}
\end{table}

% -----------------------------------------------------------------------
\section{Spectral Artefact and Associative Memory}
\label{sec:memory}
% -----------------------------------------------------------------------

After training, the model operates in a two-phase regime.

\subsection{Phase 1: Spectral Artefact Extraction}
\label{sec:memory:artefact}

At the prior mean ($\mathbf{z}=\mathbf{0}$) the decoder emits a
\emph{spectral artefact}:

\begin{equation}
  \mathcal{A}(\mathcal{I}) = \Bigl\{\hat{\Wmat},\;\hat{\boldsymbol{\omega}},\;
    \Smat_{\mathrm{mem}}\Bigr\},
  \quad
  \Smat_{\mathrm{mem}} = \sum_{k=1}^{q}\hat{\omega}_k\,
  d_\theta(\hat{\mathbf{w}}_k)\,\hat{\mathbf{w}}_k\T,
  \label{eq:artefact}
\end{equation}

where $\hat{\mathbf{w}}_k$ is the $k$-th column of $\hat{\Wmat}$ and
$d_\theta(\hat{\mathbf{w}}_k)$ is the decoder response at spectral direction $k$.
$\Smat_{\mathrm{mem}}$ is a Hopfield outer-product memory matrix~\cite{hopfield1982}.

\subsection{Phase 2: Spectral Memory Transformer}
\label{sec:memory:transformer}

The transformer feed-forward and cross-attention value matrices are
initialised from $\Smat_{\mathrm{mem}}$, providing \emph{long-term spectral
prior memory}.  Online delta-rule updates
$\Smat_{\mathrm{mem}} \leftarrow \Smat_{\mathrm{mem}} + \eta\,(\mathbf{v}-\Smat_{\mathrm{mem}}\mathbf{k})\mathbf{k}\T$
write new associations without corrupting spectral key orthogonality.
Key orthogonality (inherited from Laplacian eigenvectors) maximises
retrieval SNR by construction.

\subsection{Entropy-Controlled Generation}
\label{sec:memory:generation}

The spectral entropy of the generated Laplacian:

\begin{equation}
  H(\Lmat(\mathbf{z})) = -\sum_{k=1}^{N}\bar{\lambda}_k\log\bar{\lambda}_k,
  \quad \bar{\lambda}_k = \frac{\lambda_k}{\sum_j\lambda_j},
  \label{eq:entropy}
\end{equation}

can be steered by conditioning $\mathbf{z}$ on a target entropy level,
enabling \emph{entropy-controlled spectral graph generation}---a capability
absent in flat-prior VAEs.

% -----------------------------------------------------------------------
\section{Stability Analysis}
\label{sec:stability}
% -----------------------------------------------------------------------

\subsection{Six-Level Hierarchy}
\label{sec:stability:hierarchy}

Training stability is analysed through a six-level hierarchy~\cite{moriondo2026vdt};
failures at lower levels must be resolved before diagnosing higher ones:

\begin{enumerate}
  \item \textbf{Graph geometry.}
        $\lambda_{\max}(\Lmat_f)$ finite; CFL condition satisfied;
        graph connectivity.
  \item \textbf{Wave update per mode.}
        Damping coefficients $\gamma_k > 0$; underdamped/overdamped
        classification.
  \item \textbf{Recurrent VDT dynamics.}
        Energy $\mathcal{E}_t$ monotone; spectral entropy $H_{\text{spectral}}$
        stable.
  \item \textbf{Density matrix.}
        $\boldsymbol{\rho}_t^+$ positive semi-definite;
        $\|\boldsymbol{\rho}_t\|_F$ bounded.
  \item \textbf{Optimiser.}
        Preconditioned gradient descent convergence:
        $\kappa(\mathbf{P}_{\sigma,M}) \ll \kappa(\Lmat_f)$;
        step size $\eta < 2/L_{\sigma,M}$.
  \item \textbf{Mode weights.}
        Active mode count $N_{\text{active}} = |\{k:\E{\omega_k}>\delta\}| \geq q_{\min}$;
        $\|\hat{\Wmat}\|_F$ non-degenerate.
\end{enumerate}

\subsection{Mode Weight Collapse}
\label{sec:stability:collapse}

For high-frequency modes with large $\tau\lambda_k$, the Exponential prior pushes
$q(\omega_k)$ towards a point mass near zero.  This is \emph{correct behaviour}
(mode selection), but if it extends to low-frequency modes the loading matrix
degenerates.  Two mitigations are applied:

\begin{itemize}
  \item \textbf{Shape-parameter floor}: $a_k \geq a_{\min} = 0.1$ prevents
        full collapse.
  \item \textbf{Active-mode penalty}: a soft Lagrange term
        $\nu\cdot\mathrm{relu}(q_{\min} - N_{\text{active}})$ maintains
        a minimum number of active modes.
\end{itemize}

\subsection{Spectral Eigenvalue Floor}
\label{sec:stability:floor}

When the Fiedler eigenvalue $\lambda_2 \approx 0$ (nearly disconnected graph)
the spectral-basis KL prior becomes nearly flat and produces high gradient
variance.  A floor applied in the KL computation only
($\lambda_k \leftarrow \max(\lambda_k, 10^{-3})$) prevents numerical
instability without affecting the spectral geometry.

% -----------------------------------------------------------------------
\section{Experiments}
\label{sec:experiments}
% -----------------------------------------------------------------------

\subsection{Datasets}
\label{sec:experiments:data}

We evaluate on three benchmarks:

\begin{itemize}
  \item \textbf{CORA}~\cite{kipf2016}: 2708 nodes, 5429 edges, 7 classes,
        1433-dim bag-of-words features.
  \item \textbf{PubMed}~\cite{kipf2016}: 19717 nodes, 44338 edges, 3 classes.
  \item \textbf{Synthetic spring networks}: 400 randomly generated spring
        graphs with controllable spectral entropy.
\end{itemize}

\subsection{Evaluation Metrics}
\label{sec:experiments:metrics}

\begin{table}[ht]
  \centering
  \caption{VDT benchmark metrics.}
  \label{tab:metrics}
  \begin{tabular}{lll}
    \toprule
    \textbf{Metric} & \textbf{Symbol} & \textbf{Measures} \\
    \midrule
    Reconstruction MSE & $\|\mathbf{x}-\hat{\mathbf{x}}\|^2$ & Wiring path fidelity \\
    Latent KL & $\mathrm{KL}_z$ & Isotropic regularisation \\
    Spectral basis KL & $\mathrm{KL}_S$ & Alignment with index $\mathcal{I}$ \\
    Tau-mode KL & $\mathrm{KL}_\tau$ & Frequency band selection \\
    Active modes & $N_{\mathrm{active}}$ & Modes with $\E{\omega_k}>0.01$ \\
    Memory SNR & $d_k / N_{\text{stored}}$ & Hopfield retrieval quality \\
    ELBO Bayes factor & $\exp(\mathcal{L}(\mathcal{I}_1)-\mathcal{L}(\mathcal{I}_2))$ & Index comparison \\
    Linear probe acc. & $\mathrm{Acc}_{\text{probe}}$ & Latent discriminability \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Flagship Demo: Entropy-Controlled Spectral Generation}
\label{sec:experiments:demo}

Standard VAEs decode $\mathbf{z}$ into flat feature vectors; the VDT decodes
$\mathbf{z}$ into a \emph{graph wiring}---a Laplacian whose eigenvalues are
vibrational modes of the spring network (cf.\ Rayleigh~\cite{rayleigh1877}).
The latent space directly encodes spectral geometry, enabling samples whose
spectral entropy matches a target level.  This capability is unique to the
VDT's vibrational latent representation and is unavailable in flat-prior models.

% -----------------------------------------------------------------------
\section{Related Work}
\label{sec:related}
% -----------------------------------------------------------------------

\paragraph{Spectral graph autoencoders.}
Graph variational autoencoders~\cite{kipf2016} encode topology in the latent
space but do not use spectral priors.  VGAE and its successors decode
adjacency matrices directly, without a vibrational interpretation.

\paragraph{Probabilistic PCA on graphs.}
Laplacian Eigenmaps~\cite{belkin2003} and Diffusion Maps~\cite{coifman2006}
use the heat kernel for dimensionality reduction but are deterministic and
lack variational objectives.

\paragraph{NVIB.}
The Nonparametric Variational Information Bottleneck~\cite{nvib2022} applies
variational compression to sequence models.  The VDT shares the ELBO
philosophy but applies spectral-structural priors rather than capacity
constraints on a fixed-dimension latent.

\paragraph{Real quantum mechanics.}
Renou et al.~\cite{renou2021} and McKague et al.~\cite{mckague2009} provide
experimental and theoretical grounding for real-valued quantum theories.
The VDT does not claim to \emph{be} a quantum system, but borrows the
structure of signed-amplitude interference as a computational primitive.

% -----------------------------------------------------------------------
\section{Discussion and Future Work}
\label{sec:conclusion}
% -----------------------------------------------------------------------

The VDT establishes a formal connection between three bodies of theory:
Rayleigh's vibrational mechanics, Aaronson's real-amplitude quantum
probabilities, and Bayesian graph-spectral learning.  The central
contribution is the three-term ELBO~\eqref{eq:elbo}, which replaces hard
spectral penalties with proper variational counterparts while preserving the
ArrowSpace wiring geometry.

Several directions remain open:

\begin{itemize}
  \item \textbf{Density-matrix term (issue \#29).}
        A fourth KL over the density matrix $\boldsymbol{\rho}^+_t$
        (positive-negative amplitude structure) would complete the quantum
        analogy but is not yet in the ELBO.
  \item \textbf{VDT-Diffusion.}
        Replacing the VAE reparameterisation with a diffusion process over
        the Laplacian eigenbasis.
  \item \textbf{Scaling.}
        Large-scale evaluation on Open Graph Benchmark datasets~\cite{hu2020}.
  \item \textbf{Interactive visualisation.}
        Pluot-based~\cite{pluot} interactive exploration of spectral latent
        space is implemented in
        \texttt{demos/visualise\_spectral\_demo.py}.
\end{itemize}

% -----------------------------------------------------------------------
\section*{Acknowledgements}
% -----------------------------------------------------------------------

The ArrowSpace computational library and the VDT codebase are developed and
maintained by the tuned-org-uk team.
The architecture inherits the PPCA-to-VAE progression formalised in
\emph{The Little Book of Generative AI Foundations}~\cite{chen2026}.

% -----------------------------------------------------------------------
% Bibliography
% -----------------------------------------------------------------------
\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{rayleigh1877}
Lord Rayleigh (J.W.S.\ Strutt).
\newblock \emph{The Theory of Sound}, Volumes I and II.
\newblock Macmillan, London, 1877/1894.

\bibitem{aaronson2013}
S.\ Aaronson.
\newblock \emph{Quantum Computing Since Democritus}, Chapter~9: Quantum.
\newblock Cambridge University Press, 2013.
\newblock \url{https://scottaaronson.com/democritus/lec9.html}

\bibitem{moriondo2026arrowspace}
D.\ Moriondo.
\newblock ArrowSpace: Graph Wiring and the Differentiable Laplacian.
\newblock Technical report, tuned-org-uk, 2026.
\newblock \url{https://github.com/tuned-org-uk/pyarrowspace}

\bibitem{moriondo2026vdt}
D.\ Moriondo.
\newblock Vibrational Deduction Transformer (VDT): Spectral-PPCA Wiring
  Autoencoder with Associative Memory.
\newblock \emph{arXiv preprint}, 2026.
\newblock \url{https://github.com/tuned-org-uk/vibrational-deduction-transfomer}

\bibitem{chen2026}
T.\ Chen.
\newblock \emph{The Little Book of Generative AI Foundations}.
\newblock 2026.

\bibitem{chung1997}
F.R.K.\ Chung.
\newblock \emph{Spectral Graph Theory}.
\newblock AMS CBMS Monographs, 1997.
\newblock \url{https://mathweb.ucsd.edu/~fan/research/revised.html}

\bibitem{tipping1999}
M.E.\ Tipping and C.M.\ Bishop.
\newblock Probabilistic principal component analysis.
\newblock \emph{Journal of the Royal Statistical Society B}, 61(3):611--622, 1999.

\bibitem{kingma2013}
D.P.\ Kingma and M.\ Welling.
\newblock Auto-encoding variational Bayes.
\newblock \emph{arXiv:1312.6114}, 2013.
\newblock \url{https://arxiv.org/abs/1312.6114}

\bibitem{higgins2017}
I.\ Higgins et al.
\newblock $\beta$-VAE: Learning basic visual concepts with a constrained
  variational framework.
\newblock \emph{ICLR}, 2017.

\bibitem{belkin2003}
M.\ Belkin and P.\ Niyogi.
\newblock Laplacian Eigenmaps for dimensionality reduction and data
  representation.
\newblock \emph{Neural Computation}, 15(6):1373--1396, 2003.

\bibitem{coifman2006}
R.R.\ Coifman and S.\ Lafon.
\newblock Diffusion maps.
\newblock \emph{Applied and Computational Harmonic Analysis}, 21(1):5--30, 2006.

\bibitem{kipf2016}
T.N.\ Kipf and M.\ Welling.
\newblock Semi-supervised classification with graph convolutional networks.
\newblock \emph{arXiv:1609.02907}, 2016.

\bibitem{renou2021}
M.-O.\ Renou et al.
\newblock Quantum theory based on real numbers can be experimentally falsified.
\newblock \emph{Nature}, 600:625--629, 2021.

\bibitem{mckague2009}
M.\ McKague et al.
\newblock Simulating quantum systems using real Hilbert spaces.
\newblock \emph{Physical Review Letters}, 102:020505, 2009.

\bibitem{nvib2022}
K.\ Dib et al.
\newblock NVIB: Nonparametric variational information bottleneck.
\newblock \emph{arXiv}, 2022.

\bibitem{hopfield1982}
J.J.\ Hopfield.
\newblock Neural networks and physical systems with emergent collective
  computational abilities.
\newblock \emph{PNAS}, 79(8):2554--2558, 1982.

\bibitem{einstein1905}
A.\ Einstein.
\newblock {\"U}ber die von der molekularkinetischen Theorie der W{\"a}rme
  geforderte Bewegung von in ruhenden Fl{\"u}ssigkeiten suspendierten Teilchen.
\newblock \emph{Annalen der Physik}, 322(8):549--560, 1905.

\bibitem{jaynes1957}
E.T.\ Jaynes.
\newblock Information theory and statistical mechanics.
\newblock \emph{Physical Review}, 106(4):620--630, 1957.

\bibitem{hu2020}
W.\ Hu et al.
\newblock Open Graph Benchmark: Datasets for machine learning on graphs.
\newblock \emph{NeurIPS}, 2020.

\bibitem{pluot}
M.\ Keller.
\newblock Pluot: interactive browser-based visualisation.
\newblock \url{https://github.com/keller-mark/pluot}

\end{thebibliography}

\end{document}
